---
title: "Hidden Markov Modeling Tutorial"
author: "Marie Gilbertson"
date: "2023-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
   
   
### Preamble
What this tutorial does:

1. Demonstrates a simple example of fitting hidden Markov models (HMM) to animal movement data. 


What this tutorial does NOT do:

1. Teach you the underlying theory for HMMs, explore the vast range of uses of HMMs, or teach how to troubleshoot inital parameter selection. For that kind of information, may I suggest checking out the very thorough [guide to using the R package momentuHMM](https://cran.r-project.org/web/packages/momentuHMM/vignettes/momentuHMM.pdf). This tutorial is meant as a technical and very simple introduction to fitting HMMs with momentuHMM. 


Now that we've got the disclaimers out of the way, let's get started!  

First things first, let's load the R packages we'll need.

```{r, results='hide', message=FALSE, warning=FALSE}
##### Clear Environment #####
remove(list=ls())


#### load libraries ####
library(momentuHMM)
library(ggplot2)
library(adehabitatHR)
library(lubridate)
```
  

### Simulate data
To make it possible for anyone to replicate this workflow, we're going to simulate some animal movement data.  

We are going to be working through an example of using HMMs to identify different "behavioral states" for an animal, based on their movement data. That means that our simulated individual needs to have mutliple behavioral states. To keep things simple, we'll simulate a **resident state** characterized by short steps, meandering turning angles, and attraction to a central home range. Our other behavioral state with be a **traveling state** where our animal takes longer steps with restriced turning angles (we'll have them generally move sort of straight ahead).  

To do these simulations, we can just use a simple biased correlated random walk (BCRW) movement model, but we will need a few functions to make that happen. The R functions we'll use for this originally come from this lovely paper from [Long et al 2014](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2656.12198).  


*Note: if you're not comfortable with how we write functions in R, don't sweat it. One of these days I'll make a little tutorial about writing functions. In the meantime, just trust me that this works and that you don't need this to analyze your own real data.*

```{r simulate-data-functions}

## first we have a function for simulating a biased correlated random walk
BCRW_sim <- function(
    n=100,          #number of movement steps
    h=0.25,         #step length parameter
    rho=0,          #bias correlation parameter (0-1, where 0 -> unbiased, uncorrelated random walk, and 1 -> biased, deterministic movement)
    b=1,            #bias strength parameter
    c=0,            #bias distance decay parameter
    y0=c(0,0),       #animal starting location
    ya=c(0,0)        #animal attraction location
){
  
  #---- Main Function ------
  y <- y0
  y.t <- y
  theta.y <- runif(1,-pi,pi)       #first direction is random
  
  
  for (i in 1:n){
    
    delta <- sqrt(sum((ya-y)^2))                              #distance to attraction point 
    psi <- atan2(ya[2]-y[2],ya[1]-y[1])                       #angle toward attraction point
    beta <- tanh(b*delta^c)                                   #bias effect     
    mu <- w.circ.mean(c(theta.y,psi),c(1-beta,beta))          #biased direction
    theta.y <- rwrpnorm(1,mu,rho)                             #"draw" actual turning angle based on "expected" angle, constrained by bias correlation parameter
    #step length from chi-squared distribution
    d.y <- h*rchi(1)
    y. <- y + c(d.y*cos(theta.y),d.y*sin(theta.y))            #calculate this "step"
    
    
    #Build the trajectory
    y.t <- rbind(y.t,y.)        
    #Save for next step!
    y <- y.
  }
  
  y.out <- data.frame(y.t,row.names=NULL)
  colnames(y.out) <- c("x","y")
  
  #add date/time to trajectory; considers date/time to be on a per-minute basis
  date <- seq(1, 60*(n+1), 60)
  y.out$date <- as_datetime(date)
  return(y.out)
}

#Weighted circular mean calculation (for the bias direction)
w.circ.mean <- function (x,w) 
{
  sinr <- sum(w*sin(x)) 
  cosr <- sum(w*cos(x)) 
  circmean <- atan2(sinr, cosr) 
  circmean 
}


```
   
   
For our resident state, we'll have our animal's movements be biased towards its first location when it started in this movement state; this will cause them to stick around that general location. For our traveling state, we'll set the bias to some distant point so they head off in some other direction. We'll otherwise use the step length and bias parameters to create our two behavioral states. We'll also have our animal switch back and forth between the states a few times.  

```{r simulate-data}
set.seed(32794)



#### Simulate Dispersal ####
## We'll simulate dispersal in two "phases"
## Phase 1: resident around point (0,0)
## Phase 2: traveling toward point (50, 50)
## Phase 3: resident around new range
## Phase 4: traveling toward point (150, 100)
## Phase 5: resident around new range


# set duration (in number of locations/fixes) for phase 1
phase1.n <- 100
# simulate and view phase 1 movement
p1 <- BCRW_sim(n = phase1.n, h = 2, rho = 0.8, y0 = c(0,0), ya = c(0,0))
ggplot(p1, aes(x = x, y = y)) + geom_path() + 
  coord_fixed() + theme_bw()

# take the final location from phase 1 to use as the starting location for phase 2
end1 <- tail(p1[,c("x", "y")],1)
# simulate and view phase 2 movement
phase2.n <- 30
p2 <- BCRW_sim(n = phase2.n, h = 5, rho = 0.85, y0 = c(end1$x, end1$y), ya = c(500, 500))
ggplot(p2, aes(x = x, y = y)) + geom_path() + 
  coord_fixed() + theme_bw()

end2 <- tail(p2[,c("x", "y")],1)
# simulate and view phase 3 movement
phase3.n <- 100
p3 <- BCRW_sim(n = phase3.n, h = 2, rho = 0.8, y0 = c(end2$x, end2$y), ya = c(end2$x, end2$y))
ggplot(p3, aes(x = x, y = y)) + geom_path() + 
  coord_fixed() + theme_bw()


# take the final location from phase 3 to use as the starting location for phase 4
end3 <- tail(p3[,c("x", "y")],1)
# simulate and view phase 2 movement
phase4.n <- 30
p4 <- BCRW_sim(n = phase4.n, h = 5, rho = 0.85, y0 = c(end3$x, end3$y), ya = c(1000, 0))
ggplot(p4, aes(x = x, y = y)) + geom_path() + 
  coord_fixed() + theme_bw()


end4 <- tail(p4[,c("x", "y")],1)
# simulate and view phase 3 movement
phase5.n <- 100
p5 <- BCRW_sim(n = phase5.n, h = 2, rho = 0.8, y0 = c(end4$x, end4$y), ya = c(end4$x, end4$y))
ggplot(p5, aes(x = x, y = y)) + geom_path() + 
  coord_fixed() + theme_bw()



# remove last step of each phase trajectory so no duplicated locations when all phases are combined...except sometimes GPS collars accidentally transmit the same location twice (which can be a problem in HMMs), so let's leave a couple duplicate locations for use to "clean up" later. 
p1 <- p1[-c(nrow(p1)),] 
p3 <- p3[-c(nrow(p3)),] 



# combine phases and view
p <- rbind(p1, p2, p3, p4, p5)
ggplot(p, aes(x = x, y = y)) + geom_path() + 
  coord_fixed() + theme_bw()

# add time stamps; we'll say that we have locations every 4 hours
p$date <- seq(as.POSIXct("2017-04-01 04:00:00"), by = (4*60*60), length.out = nrow(p))


```


To be more like real telemetry data, let's add some noise to our time stamps. GPS collars really never collect a location at perfect intervals, and this can impact our HMM, so let's allow all our time stamps to randomly vary by up to 10 minutes before or after the "scheduled," every-4-hours time.
```{r fix-rate-noise}
p$real.time <- p$date + sample(c((-10*60):(10*60)), size = nrow(p), replace = T)
head(p)
```

In reality, we'd also be missing some locations because, for example, the collar would fail to transmit the animal's location. These missing locations can impact our HMM, so let's also randomly drop 2% of our locations. 
```{r drop-locs}
drop <- sample(1:nrow(p), size = 0.02*nrow(p), replace = F)
pd <- p[-c(drop),]

ggplot(p, aes(x = x, y = y, color = real.time)) + geom_path() + 
  coord_fixed() + theme_bw() + ggtitle("Simulated 'messy' data")
```

Now we have a nice simulated dataset with a little added "messiness" to replicate what you might experience with real animal movement data. Naturally, that means it's time to "clean up" our data!  

### Prepare animal movement data

What's the point in making our data "messy" if we're just going to go and clean it up, you ask? Well, if you're working with real data, it's going to messy and it does no one any good to pretend otherwise. 

To begin with, let's explore our data a little bit, pretending that we didn't just simulate this ourselves. We can start by making our animal movement data an "ltraj" object so we can use some of the functions in the very handy *adehabitatLT* package. 

```{r make-ltraj}

traj <- as.ltraj(xy = pd[,c("x", "y")], date = pd$real.time, id = "ind.1")
traj
```

We can see that we now have an "ltraj" object containing the movement data for "ind.1." There are 353 relocations, but since we haven't added any "NAs" to our data, our object supposedly doesn't have missing data (but you and I know that's not true). We can also see the start and end dates for our data. Lastly, we can also see that we have both locations and time stamps (ltraj considers these "Type II" trajectories), and that there is a variable time lag between relocations.  

An ltraj object is fundamentally a list of animal movement trajectories. If we want to view the trajectory data itself, we can access it using what I'll call normal R list syntax. In other words, if we want to look at the first list item, we use double brackets to do so. In this case, when we look at the data itself, we see that converting our simulated data to an ltraj object has calculated lots of useful things like step lengths, turning angles, etc. 
```{r view-traj}
head(traj[[1]])
```
  
We can then plot lots of useful things! For example, we can look at the trajectory itself, or things like the sampling frequency/fix rate over time. In addition, we can look at "net squared displacement" (this column is called "R2n" in the ltraj object); this shows an animal's travel or displacement from its starting location, which can be handy for spotting shifts to new ranges.

```{r}
# plot the trajectory
plot(traj)

# plot sampling frequency (note how you can see some variability in the fix rate AND that we're missing locations!)
plotltr(traj, "dt/3600") # dt = change in time between locations, and since it's in seconds, we'll divide by 3600 to view time in hours

# plot net squared displacement
plotltr(traj, "R2n") # we can see that this individual had three main ranges, with a couple of shifts in between them
```
  
Now that we've taken a look at our data, to fit HMMs, we need to make our trajectory **regular**: this means we need equal time steps between locations. We can fix some of the irregularity by just rounding the timing of our fixes to the nearest hour, since our data was "collected" at the top of the hour every 4 hours. If our data were collected every 15 minutes, we could simply round to that fix schedule. Let's start by doing this rounding procedure:

```{r round-fixes}
#### prep the data ####
### need to make trajectory regular (except for missing locations) by rounding timing of locations ###

# convert trajectory object to a dataframe
ddf <- ld(disp)
head(ddf)

# update print.POSIXct so that it always gives HH:MM:SS (otherwise, drops times for midnight)
print.POSIXct <- function(x,...)print(format(x,"%Y-%m-%d %H:%M:%S"))

# function for rounding date/time to nearest hour (returns error if input is greater than 15 minutes from nearest hour)
round.datetime <- function(x){
  minute <- as.numeric(format(x, "%M"))
  if(minute<30){
    out <- as.POSIXct(paste0(format(x, "%Y-%m-%d %H"), ":00:00"), tz = "America/Chicago")
  }else if(minute>=30){
    hour <- as.numeric(format(x, "%H"))+1
    
    if(hour==25){
      hour <- 1
      day <- as.Date(x, tz = "America/Chicago")+1
      out <- as.POSIXct(paste0(day, " ", hour, ":00:00"), tz = "America/Chicago")
    }else{
      out <- as.POSIXct(paste0(format(x, "%Y-%m-%d"), " ", hour, ":00:00"), tz = "America/Chicago")
    }
  }
  
  dt <- abs(difftime(x, out, tz = "America/Chicago", units = "mins"))
  
  if(dt > 15){
    out <- "error"
  }else{
    out <- format(out, "%Y-%m-%d %H:%M:%S") # need to output as character string and convert back to POSIXct later
  }
  return(out)
}


# use new function to round date/times and view number of fixes that fell more than 15 min from top of the hour
ddf$round.time <- sapply(ddf$date, round.datetime)
nrow(ddf[ddf$round.time=="error",])

# keep the fixes that are within 15 min of the top of the hour
ddf2 <- ddf[ddf$round.time!="error",]
ddf2$round.time <- as.POSIXct(ddf2$round.time, tz = "America/Chicago")

# check frequency of fixes, regularity of data
ddf2$dt2 <- NA
for(i in 2:nrow(ddf2)){
  ddf2$dt2[i] <- difftime(ddf2$round.time[i], ddf2$round.time[i-1], tz = "America/Chicago", units = "hours")
}
table(ddf2$dt2)


# drop the fixes that don't have a unique date/time stamp (time difference from previous fix = 0)
ddf3 <- subset(ddf2, ddf2$dt2 != 0 | is.na(ddf2$dt2))

```


```{r}

# drop the once daily fixes at the end of the trajectory (dispersal has already occurred and don't want to interpolate across those big gaps)
ddf3 <- subset(ddf3, as.Date(ddf3$round.time, "America/Chicago") < as.Date("2017-11-01", tz = "America/Chicago"))
table(ddf3$dt2) # should only have time differences of 1, 2, 3, 4, or 8 hours now
# view the time span, frequency of fixes
plot(ddf3$round.time, rep("1", nrow(ddf3)))

# move forward with this regularized, trimmed dataset
ddf <- ddf3


```